---
title: "Fertility demo"
author: "Amy Anderson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Known Unknowns: Replicating the Main Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, 
                      
  dpi     = 300,      # crisp resolution
  fig.width  = 10,    # wide enough for two side-by-side plots
  fig.height = 5,     # adjust to taste
  out.width  = "100%" # fills the vignette column width
)

```

### Persephone has joined forces with Demeter!

The mortality model you know and love now has fertility built into its population simulation engine. 



```{r Load libraries}
library(persephone)
library(tidyverse)
library(ggplot2)
library(patchwork)
```
 
 
 
### Step 1: Generating Age-structured starting populations  
None of the fertility functions will run in a simulation with an age cohort, in which all agents are the same age. The starting population must be age-structured in order for fertility rate to be specified.  
```{r, fig.cap="Age structure of initial agents is determined by mortality regime and population growth rate, \n which defaults to r = 0 (a stationary population)"}

# Hold population configuration parameters here for ease of reference in functions that follow. 
  pop_config <- list(
    model_lesions              = FALSE,
    annual_exposure            = NULL,
    model_frailty              = FALSE,
    gammafrailty_shape         = NULL,
    # nothing below here should get called by the fertility functions. 
    gammafrailty_scale         = NULL,
    lesion_formation_window    = c(0,0),
    exposure_causes_hazard     = FALSE,
    hazard_is_transient        = FALSE,
    lesion_requires_survival   = FALSE,
    exposure_hazard_multiplier = 1
  )


# create a cohort of agents, aged 0.
cohort <- create_pop(pop0_size = 100, 
                      age_structured = FALSE,
                      pop_config = pop_config,
                      mortality_regime = CoaleDemenyWest5)

# Generate age-structured, stationary populations of 10,000 people.
pop_cdw3 <- create_pop(pop0_size = 10000, 
                      age_structured = TRUE,
                      pop_config = pop_config,
                      r = 0, # population growth rate
                      mortality_regime = CoaleDemenyWestF3)  # age structure is pushed by the mortality regime. 
  

pop_cdw21 <- create_pop(pop0_size = 10000, 
                      age_structured = TRUE,
                      pop_config = list(model_lesions = FALSE, 
                                        model_frailty = FALSE),
                      r = 0, # population growth rate
                      mortality_regime = CoaleDemenyWestF21) 

# plot the age distributions created in the two stationary populations:
p1 <- ggplot(pop_cdw3, aes(x = age)) +
  geom_bar(color = "blue") +
  labs(title = "CDW F Level 3", subtitle ="stationary population, n = 10,000") +
 # scale_y_continuous(limits = c(0,700)) +
  theme_bw()
p2 <- ggplot(pop_cdw21, aes(x = age)) +
  geom_bar() + 
    labs(title = "CDW F Level 21", subtitle = "stationary population, n = 10,000") +
   # scale_y_continuous(limits = c(0,700)) +
  theme_bw()

p1 + p2
```
                      
                      
\n
\n
Now make them stable, but not stationary:
```{r, fig.cap="Holding mortality constant and not modeling migration, population growth/decline is assumed to be entirely a function of fertility in these three plots. The huge impact of fertility rates on the age structure of a mortality sample is apparent here."}
pop_cdw3_growing <- create_pop(pop0_size = 10000, 
                      age_structured = TRUE,
                      pop_config = list(model_lesions = FALSE, 
                                        model_frailty = FALSE),
                      r = 0.03, # population growth rate
                      mortality_regime = CoaleDemenyWestF3) 

pop_cdw3_declining <- create_pop(pop0_size = 10000, 
                      age_structured = TRUE,
                      pop_config = list(model_lesions = FALSE, 
                                        model_frailty = FALSE),
                      r = -0.03, # population growth rate
                      mortality_regime = CoaleDemenyWestF3)

p1 <- p1 + 
    scale_y_continuous(limits = c(0,1500)) +
  labs(title = "stationary population", subtitle = "Mortality: CDW F Level 3")

p3 <- ggplot(pop_cdw3_growing, aes(x = age)) +
  geom_bar(color = "blue") +
  labs(subtitle = "Mortality: CDW F Level 3", title ="3% population growth") +
  scale_y_continuous(limits = c(0,1500)) +
  theme_bw()

p4 <- ggplot(pop_cdw3_declining, aes(x = age)) +
  geom_bar(color = "blue") +
  labs(subtitle = "Mortality: CDW F Level 3", title ="3% population decline") +
  scale_y_continuous(limits = c(0,1500)) +
  theme_bw()

p4 + p1 + p3
```
\n
\n

### Step 2: new-born agents every year 
the call for `Simulate_Cemetery` now includes an argument called `tfr`: average Total Fertility Rate for the population. 
It gets called in the time step loop of the Simulate_Cemetery ABM by the function 'apply_fertility'. 

```{r, fig.cap="Age-specific probability of giving birth. In combination with TFR and the existing age structure of the population, this determines the expected number of births each year."}
# example of the asfr trapezoid
birth_probs <- compute_trapezoid_asfr(ages = unique(pop_cdw3$age), tfr = 4)
asfr_range <- data.frame(ages = unique(pop_cdw3$age), birth_probs)
ggplot(asfr_range, aes(x = ages, y = birth_probs)) +
  geom_line(linetype = "dashed") +
  labs(x = "Age", y = "Probability of giving birth") +
  theme_bw()
```


\n
\n

So let's take our stationary CDW3 population and apply one year of fertility probability to them: 
```{r, fig.cap="After applying the fertility function once, the number of infants is higher. No other ages should be affected by this transformation."}

# Let's say each woman has, on average, 4 children. 
tfr <- 4

# asfr = age-structured fertility rate.
# (this is called inside the main ABM loop if age_structured == TRUE)
asfr <- compute_trapezoid_asfr(ages = 0:100, tfr = tfr)

updated_pop <- apply_fertility(pop_cdw3, # the current population
                               tfr = tfr,
                               dx = 1,
                               time = 1, # the current year. Could be any integer. 
                               asfr = asfr, # the age-structured fertility rate
                               config = pop_config)

updated_pop$fertility_applied <- "After new births"    
pop_cdw3$fertility_applied <- "Before new births"
before_and_after <- rbind(updated_pop, pop_cdw3)

ggplot(before_and_after, aes(x = age)) +
  geom_bar(aes(fill = fertility_applied), position = "dodge") +
  labs(x = "Age", y = "n", fill = " ") +
  theme_bw()
```
\n
\n
\n
... Now let's run a stable population through time.
```{r, fig.cap="Variation in population growth across 300 years when mortality is held constant and TFR ranges from 0 to 6.", echo=FALSE}
params <- get_default_params()
params$pop0_size = 100
params$max_years = 300
params$age_structured = TRUE
params$mortality_regime = CoaleDemenyWestF11


test_tfr <- c(0, 1, 2.5, 3.5, 5, 6)
test_pops <- vector("list", length(test_tfr))
pop_size <- vector("list", length(test_tfr))
plots <- vector("list", length(test_tfr))
for(i in 1:length(test_tfr)){
  params$tfr = test_tfr[i]
  test_pops[[i]] <- do.call(Simulate_Cemetery, params)
  pop_size[[i]] <- test_pops[[i]]$annual_census
  pop_size[[i]]$tfr <- test_tfr[i]
   plots[[i]] <- ggplot(test_pops[[i]]$annual_census, aes(x = Time, y = n)) +
   geom_line() +
     ggtitle(paste0("TFR = ", test_tfr[i])) +
   theme_bw()
}

# collect all population censuses into a single data frame, with TFR values
# tfr_sweep_n <- do.call(rbind, pop_size)
# # plot population size across time, with color mapped to TFR
# ggplot(tfr_sweep_n, aes(x = Time, y = n)) +
#   geom_line(aes(group = tfr, color = as.factor(tfr))) +
#   labs(x = "Year", y = "Population size", color = "TFR") +
#   theme_bw()
# Plotting in a single panel only works for small populations on short time frames, when the range of TFRs isn't too wide. 


# Compare annual population size over a 300-year simulation run when TFR ranges between 0 and 6 and mortality stays stable. 
plots[[1]] + plots[[2]] + plots[[3]] + plots[[4]]  + plots[[5]]  + plots[[6]] + plot_layout(ncol = 3)
```

\n
\n

...and now, something approximating an agrarian population.  
```{r Imaginary Agrarians, fig.cap="Imaginary Agrarians. small starting population (n = 100), high mortality (CDW3), high fertility (TFR = 12).", warning=FALSE}

params$mortality_regime <- CoaleDemenyWestF3
params$tfr <- 12
params$max_years <- 200
params$pop0_size <- 100

agrarian <- do.call(Simulate_Cemetery, params)

View(agrarian$individual_outcomes)

n_alive <- ggplot(agrarian$annual_census, aes(x = Time, y = n)) +
  geom_line() +
  labs(x = "Year", y = "Population Size") +
  theme_bw()

n_born <- agrarian$individual_outcomes %>%
  group_by(year_born) %>%
  summarise(n_born = n()) %>%
  ggplot(aes(x = year_born, y = n_born)) +
  geom_line() +
  labs(x = "Year born", y = "Number of Births") +
  theme_bw()

n_died <- agrarian$individual_outcomes %>%
  group_by(year_died) %>%
  summarise(n_died = n()) %>%
  ggplot(aes(x = year_died, y = n_died)) +
  geom_line() +
  labs(x = "Year died", y = "Number of Deaths") +
  theme_bw()

n_alive / (n_born + n_died)

```

