---
title: "Fertility demo"
author: "Amy Anderson"
date: "2026-06-12"
output: html_document
---

# Friday To-Do:

# 4. Write a script for fertility functions demo
# - remember to go back and change the placement of the column creation for was_deposited and in_sample
# - fix the 'annual_census' output. 
# - generate a dynamic population
# - plot the population size through time.

# 5. Can you make the package updated and downloadable? 
# ...and publish the Shiny app?



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Persephone has joined forces with Demeter!

The mortality model you know and love now has fertility built into its population simulation engine. 



```{r Load libraries}
library(persephone)
library(tidyverse)
library(ggplot2)
library(patchwork)
```
 
 
 
### Fertility update 1: Generating Age-structured starting populations
```{r}

# Hold population configuration parameters here for ease of reference in functions that follow. 
  pop_config <- list(
    model_lesions              = FALSE,
    annual_exposure            = NULL,
    model_frailty              = FALSE,
    gammafrailty_shape         = NULL,
    # nothing below here should get called by the fertility functions. 
    gammafrailty_scale         = NULL,
    lesion_formation_window    = c(0,0),
    exposure_causes_hazard     = FALSE,
    hazard_is_transient        = FALSE,
    lesion_requires_survival   = FALSE,
    exposure_hazard_multiplier = 1
  )


# create a cohort of agents, aged 0.
cohort <- create_pop(pop0_size = 100, 
                      age_structured = FALSE,
                      pop_config = pop_config,
                      mortality_regime = CoaleDemenyWest5)

# Generate age-structured, stationary populations of 10,000 people.
pop_cdw3 <- create_pop(pop0_size = 100, 
                      age_structured = TRUE,
                      pop_config = pop_config,
                      r = 0, # population growth rate
                      mortality_regime = CoaleDemenyWestF3)  # age structure is pushed by the mortality regime. 
  

pop_cdw21 <- create_pop(pop0_size = 10000, 
                      age_structured = TRUE,
                      pop_config = list(model_lesions = FALSE, 
                                        model_frailty = FALSE),
                      r = 0, # population growth rate
                      mortality_regime = CoaleDemenyWestF21) 


# plot the age distributions created in the two stationary populations:
p1 <- ggplot(pop_cdw3, aes(x = age)) +
  geom_bar(color = "blue") +
  labs(title = "CDW F Level 3", subtitle ="stationary population, n = 10,000") +
  scale_y_continuous(limits = c(0,700)) +
  theme_bw()
p2 <- ggplot(pop_cdw21, aes(x = age)) +
  geom_bar() + 
    labs(title = "CDW F Level 21", subtitle = "stationary population, n = 10,000") +
    scale_y_continuous(limits = c(0,700)) +
  theme_bw()

p1 + p2
```
                      

Now make them stable, but not stationary:
```{r}
pop_cdw3_growing <- create_pop(pop0_size = 10000, 
                      age_structured = TRUE,
                      pop_config = list(model_lesions = FALSE, 
                                        model_frailty = FALSE),
                      r = 0.03, # population growth rate
                      mortality_regime = CoaleDemenyWestF3) 

pop_cdw3_declining <- create_pop(pop0_size = 10000, 
                      age_structured = TRUE,
                      pop_config = list(model_lesions = FALSE, 
                                        model_frailty = FALSE),
                      r = -0.03, # population growth rate
                      mortality_regime = CoaleDemenyWestF3)

p1 <- p1 + 
    scale_y_continuous(limits = c(0,1500)) +
  labs(subtitle = "stationary population")

p3 <- ggplot(pop_cdw3_growing, aes(x = age)) +
  geom_bar(color = "blue") +
  labs(title = "CDW F Level 3", subtitle ="3% population growth") +
  scale_y_continuous(limits = c(0,1500)) +
  theme_bw()

p4 <- ggplot(pop_cdw3_declining, aes(x = age)) +
  geom_bar(color = "blue") +
  labs(title = "CDW F Level 3", subtitle ="3% population decline") +
  scale_y_continuous(limits = c(0,1500)) +
  theme_bw()

p4 + p1 + p3
```

### Fertility update 2: new-born agents every year 
the call for `Simulate_Cemetery` now includes an argument called `tfr`: average Total Fertility Rate for the population. 
It gets called in the time step loop of the Simulate_Cemetery ABM by the function 'apply_fertility'. 

Here's a look at apply_fertility and the functions it relies on: 
```{r Fertility functions}

#' Compute a trapezoid age-specific fertility schedule from a TFR
#'
#' The schedule rises linearly from 0 at age_start to a peak at age_peak_start,
#' remains flat to age_peak_end, then falls linearly back to 0 at age_end.
#' The peak rate is derived analytically so that the sum of ASFRs across all
#' ages equals the TFR (i.e. the schedule integrates correctly by construction).
#'
#' @param ages Integer vector of ages to compute the schedule over
#' @param tfr Numeric. Total Fertility Rate (expected lifetime births per woman)
#' @param age_start Numeric. Age at which fertility begins. Default 15.
#' @param age_peak_start Numeric. Age at which fertility reaches its peak. Default 25.
#' @param age_peak_end Numeric. Age at which fertility begins to decline. Default 35.
#' @param age_end Numeric. Age at which fertility reaches zero. Default 45.
#' @return Named numeric vector of annual age-specific fertility rates
#' @keywords internal
#' 
#' 
#.  Test how changing age start, peak, end changes other variables in the model. 
compute_trapezoid_asfr <- function(ages, tfr,
                                   age_start = 15, age_peak_start = 25,
                                   age_peak_end = 35, age_end = 45) {
  flat_width      <- age_peak_end   - age_peak_start
  ramp_up_width   <- age_peak_start - age_start
  ramp_down_width <- age_end        - age_peak_end
  area_per_peak   <- flat_width + ramp_up_width / 2 + ramp_down_width / 2
  peak_asfr       <- tfr / area_per_peak
  
  asfr <- dplyr::case_when(
    ages <  age_start      ~ 0,
    ages <  age_peak_start ~ peak_asfr * (ages - age_start)  / ramp_up_width,
    ages <= age_peak_end   ~ peak_asfr,
    ages <  age_end        ~ peak_asfr * (age_end - ages)    / ramp_down_width,
    TRUE               ~ 0
  )
  
  setNames(asfr, ages)
}

# example of the asfr trapezoid
birth_probs <- compute_trapezoid_asfr(ages = unique(pop_cdw3$age), tfr = 4)
tmp <- data.frame(ages = unique(pop_cdw3$age), birth_probs)
plot(tmp$ages, tmp$birth_probs)
```
```{r Fertility functions cont.}
#' Generate new agent rows to append to the pop
#'
#' Creates n_births new agents with age 0 and no lesion, with agent_ids
#' continuing sequentially from the current maximum in the population. All other
#' columns are initialized to match the structure of create_pop().
#'
#' @param pop Population data frame (used to determine next agent_id)
#' @param n_births Integer. Number of new agents to create.
#' @param pop_config A list of parameters for the initial population so that new agents will have matching trait columns. 
#' @return Data frame with n_births rows ready to rbind() onto the population
#' @keywords internal

generate_births <- function(pop, n_births, pop_config) {
  if (n_births == 0L) return(pop[0, ])
  
  max_id <- max(pop$agent_id)
  
  new_agents <- data.frame(
    agent_id = seq(max_id + 1, max_id + n_births),
    age      = 0
  )
  
  if (pop_config$model_lesions) {
    new_agents$lesion <- 0
  }
  if (pop_config$model_frailty) {
    new_agents$frailty          <- rgamma(n_births,
                                          shape = pop_config$gammafrailty_shape,
                                          scale = pop_config$gammafrailty_scale)
    new_agents$acquired_frailty <- NA_real_
  }
  
  if ("n_stress_events" %in% names(pop)) {
    new_agents$n_stress_events <- 0L
  }
  
  new_agents[, names(pop), drop = FALSE]
}


#' Apply fertility to the living population for one timestep
#'
#' Determines how many births occur this timestep and appends new agents to
#' the pop. The number of births is drawn from a Poisson distribution whose
#' rate equals the sum of age-specific fertility rates across all reproductive-
#' age women in the living population.
#'
#' Women are approximated as half of all living agents aged between age_start
#' and age_end (inclusive). This assumption can be replaced once sex is added
#' as a column to the pop data frame, by filtering on pop$sex == "F"
#' instead.
#'
#' Birth counts are drawn as Poisson rather than Bernoulli because multiple
#' births can occur at a given age in a population of any size, and the Poisson
#' approximation to the sum of many independent Bernoulli trials is exact in
#' the limit of large population and small per-individual rate.
#'
#' @param pop Population data frame
#' @param tfr Numeric. Total Fertility Rate.
#' @param asfr Named numeric vector of age-specific fertility rates, as
#'   returned by compute_trapezoid_asfr(). The names must be character
#'   representations of integer ages.
#' @param dx Numeric. timestep size. Scales fertility rates proportionally.
#'   Default 1 (annual timestep).
#' @param pop_config A list of population trait parameter values to be called so that newborns have the same columns as the existing population. 
#' @return Updated pop data frame with new agents appended
#' @keywords internal
#' I want apply_fertility to reference create_pop and generate sensible birth values for the same columns that are specified in the create_pop call. 
apply_fertility <- function(pop, tfr, asfr, dx = 1, pop_config) {
  
  repro_ages      <- as.numeric(names(asfr))
  in_repro_window <- pop$age %in% repro_ages
  repro_ages_actual <- pop$age[in_repro_window]
  
  n_repro         <- sum(in_repro_window)
  n_female_approx <- round(n_repro / 2)
  
  if (n_female_approx == 0L) return(pop)
  
  female_age_sample <- sample(repro_ages_actual,
                              size    = n_female_approx,
                              replace = FALSE)
  
  asfr_values     <- asfr[as.character(female_age_sample)]
  expected_births <- sum(asfr_values * dx) # <- You should change apply_mortality to follow this logic too. Right now it only works if dx = 1. 
  n_births        <- rpois(1, lambda = expected_births)
  
  new_agents <- generate_births(pop, n_births, pop_config)
  rbind(pop, new_agents)
}
```


So let's take our stationary CDW3 population and apply one year of fertility probability to them: 
```{r}
# Let's say each woman has, on average, 4 children. 
tfr <- 4

# asfr = age-structured fertility rate.
# (this is called inside the main ABM loop if age_structured == TRUE)
asfr <- compute_trapezoid_asfr(ages = 0:100, tfr = tfr)

updated_pop <- apply_fertility(pop_cdw3, # the current population
                               tfr = tfr,
                               asfr, # the age-structured fertility rate
                               pop_config = pop_config)

View(updated_pop)   

p5 <- ggplot(updated_pop, aes(x = age)) +
  geom_bar(color = "blue") +
  labs(title = "CDW F Level 3", subtitle ="TFR = 4, one year") +
  scale_y_continuous(limits = c(0,1500)) +
  theme_bw()

p1 + p5
```

... Now let's run a stable population through time.
```{r}
params <- get_default_params()
params$max_years = 500
params$age_structured = TRUE
params$mortality_regime = CoaleDemenyWestF11
params$tfr = 3.1

test_tfr <- c(1, 2, 2.8, 3.5, 5, 7)
test_pops <- vector("list", length(test_tfr))
plots <- vector("list", length(test_tfr))
for(i in 1:length(test_tfr)){
  params$tfr = tfr[i]
  test_pops[[i]] <- do.call(Simulate_Cemetery, params)
  plots[[i]] <- ggplot(test_pops[[i]]$annual_census, aes(x = Time, y = n)) +
  geom_line() +
  theme_bw()
}

# Compare annual population size over a 500-year simulation run when TFR ranges between 1 and 7 and mortality stays stable. 
plots[[1]] + plots[[2]] + plots[[3]] + plots[[4]] # + plots[[5]] # + plots[[6]] + plot_layout(ncol = 3)


params$mortality_regime <- CoaleDemenyWestF3
params$tfr <- 12
params$max_years <- 100
params$pop0_size <- 100
agrarian <- do.call(Simulate_Cemetery, params)
View(agrarian$individual_outcomes)

ggplot(agrarian$annual_census, aes(x = Time, y = n)) +
  geom_line() +
  theme_bw()

agrarian$individual_outcomes %>%
  group_by(year_died) %>%
  summarise(n_died = n()) %>%
  ggplot(aes(x = year_died, y = n_died)) +
  geom_line()


tail(agrarian$individual_outcomes %>%
  group_by(year_died) %>%
  summarise(n_died = n()))
```

